Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-11-30 tmm.tex
Add one more figure for consideration
Settle down experimental results

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best online video watching experience, an adaptive video streaming system which can adjust video quality according to actual bandwidth condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 0.63 to 0.21, and at the same time deliver higher video quality, with an average PSNR gain of 0.66dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include online video-on-demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin playing the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better streaming service, some researchers choose to adopt the new software defined networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of adaptive video manycast \cite{Xue15}. Although these studies offer a promising way to optimize network conditions, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to cellular network). Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective linear error model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its R-D impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

@@

The following of this paper will present our solution for each of the problems in detail. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system.

@@

The performance comparisons for City, Foreman, Harbour, and Soccer are shown in Fig. \ref{fig:extraction-performance}. It can be seen that the R-D curves of the proposed method are always on the top. For each sequence, the maximum and average PSNR gain through all 10 bitrate points are calculated and listed in Table \ref{tab:results}, which shows that our extraction method can achieve a significant performance gain compared with JSVM.

@@

After combining the two algorithms together, the proposed adaptive streaming system is able to deliver a higher and smoother video quality for the user. We compare the proposed system's performance with DSS that supports SVC but has its original quality control method and JSVM Basic bitstream extraction. The result for the same test stream as used in subsection \ref{subsec:exp-control} is present in Fig. \ref{fig:fluctuation-psnr}. It shows PSNR change perceived by the user when available bandwidth fluctuates around certain values (600kbps, 800kbps, 1000kbps and 1200kbps). The bandwidth conditions are simulated with NetLimiter and intense fluctuations are observed. It can be seen that, the proposed system performs better in that it has effectively avoided unnecessary quality drop (e.g., note the curve around time 450s in Fig. \ref{fig:Fluctuation-PSNR1} or 250s in Fig. \ref{fig:Fluctuation-PSNR3}) and been able to keep a relatively higher PSNR value (e.g., note the curve around time 400s in Fig. \ref{fig:Fluctuation-PSNR2} or 500s in Fig. \ref{fig:Fluctuation-PSNR4}). The average and variance value of PSNR is also calculated and present in Fig. \ref{fig:fluctuation-psnr}, showing clearly the proposed system's advantage. For instance, in Fig. \ref{fig:Fluctuation-PSNR4} where bandwidth fluctuates around 1200kbps, the average PSNR delivered by the proposed system is 41.14dB, 0.66dB higher than that of the baseline system, and the PSNR variance is also reduced from 0.63 to 0.21, indicating a more smooth video quality change.

\begin{figure*}[t]
\centering
\subfloat[Bandwidth fluctuates around 400kbps: PSNR average and variance in the original system are 36.00 and 0.30, while in the proposed system are 36.50 (increased by 0.50) and 0.23 (reduced by 0.07).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR1.png}
\label{fig:Fluctuation-PSNR1}}\hfill
\subfloat[Bandwidth fluctuates around 600kbps: PSNR average and variance in the original system are 37.06 and 0.61, while in the proposed system are 37.32 (increased by 0.26) and 0.33 (reduced by 0.28).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR2.png}
\label{fig:Fluctuation-PSNR2}}
\qquad
\subfloat[Bandwidth fluctuates around 800kbps: PSNR average and variance in the original system are 37.92 and 1.24, while in the proposed system are 38.75 (increased by 0.83) and 0.69 (reduced by 0.55).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR3.png}
\label{fig:Fluctuation-PSNR3}}\hfill
\subfloat[Bandwidth fluctuates around 1000kbps: PSNR average and variance in the original system are 38.28 and 1.30, while in the proposed system are 38.94 (increased by 0.66) and 0.71 (reduced by 0.59).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR4.png}
\label{fig:Fluctuation-PSNR4}}
\caption{User-perceived PSNR change with time when bandwidth fluctuates \label{fig:fluctuation-psnr}}
\end{figure*}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best online video watching experience, an adaptive video streaming system which can adjust video quality according to actual bandwidth condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with an average PSNR gain of 0.83dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include online Video-on-Demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin playing the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better streaming service, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of adaptive video manycast \cite{Xue15}. Although these studies offer a promising way to optimize network conditions, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to cellular network). Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective Linear Error Model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its Rate-Distortion (R-D) impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

@@

The following of this paper will present our solution for each of the problems in detail. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system. It should be noted that, in practical video streaming systems there are other issues that need to be addressed. One of them is occasional packet loss during transmission, which will cause decoding errors and corrupted images, resulting in degradation of video quality. However, in this paper we will focus on server-side adaptation and simply assume that no packet loss is occurred.

@@

The performance comparisons for City, Foreman, Harbour, and Soccer are shown in Fig. \ref{fig:extraction-performance}. It can be seen that the R-D curves of the proposed method are always on the top. For each sequence, the maximum and average PSNR gain through all 10 bitrate points are calculated and listed in Table \ref{tab:extraction-gain}, which shows that our extraction method can achieve a significant performance gain compared with JSVM.

@@

After combining the two algorithms together, the proposed adaptive streaming system is able to deliver a higher and smoother video quality for the user. We compare the proposed system's performance with DSS that supports SVC but has its original quality control method and JSVM Basic bitstream extraction. The result for the test sequence Soccer is present in Fig. \ref{fig:fluctuation-psnr}. It shows the streamed video's PSNR change when available bandwidth fluctuates around certain values (400kbps, 600kbps, 800kbps and 1000kbps). The bandwidth conditions are simulated with NetLimiter and intense fluctuations are observed. It can be seen that, the proposed system performs better in that it has effectively avoided unnecessary quality drop (e.g., note the curve around time 350s in Fig. \ref{fig:Fluctuation-PSNR3} or 150s in Fig. \ref{fig:Fluctuation-PSNR4}) and been able to keep a relatively higher PSNR value (e.g., note the curve around time 450s in Fig. \ref{fig:Fluctuation-PSNR1} or 500s in Fig. \ref{fig:Fluctuation-PSNR2}). The average and variance value of PSNR is also calculated and present in Fig. \ref{fig:fluctuation-psnr}, showing clearly the proposed system's advantage. For instance, in Fig. \ref{fig:Fluctuation-PSNR3} where bandwidth fluctuates around 800kbps, the average PSNR delivered by the proposed system is 38.75dB, 0.83dB higher than that of the baseline system, and the PSNR variance is also reduced from 1.24 to 0.69, indicating a more smooth video quality change.

\begin{figure*}[t]
\centering
\subfloat[Bandwidth fluctuates around 400kbps: PSNR average and variance in the original system are 36.00 and 0.30, while in the proposed system are 36.50 (increased by 0.50) and 0.23 (reduced by 0.07).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR1.png}
\label{fig:Fluctuation-PSNR1}}\hfill
\subfloat[Bandwidth fluctuates around 600kbps: PSNR average and variance in the original system are 37.06 and 0.61, while in the proposed system are 37.32 (increased by 0.26) and 0.33 (reduced by 0.28).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR2.png}
\label{fig:Fluctuation-PSNR2}}
\qquad
\subfloat[Bandwidth fluctuates around 800kbps: PSNR average and variance in the original system are 37.92 and 1.24, while in the proposed system are 38.75 (increased by 0.83) and 0.69 (reduced by 0.55).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR3.png}
\label{fig:Fluctuation-PSNR3}}\hfill
\subfloat[Bandwidth fluctuates around 1000kbps: PSNR average and variance in the original system are 38.28 and 1.30, while in the proposed system are 38.94 (increased by 0.66) and 0.71 (reduced by 0.59).]{
\includegraphics[width=0.47\textwidth]{Fluctuation-PSNR4.png}
\label{fig:Fluctuation-PSNR4}}
\caption{The streamed video's PSNR change with time when bandwidth fluctuates \label{fig:fluctuation-psnr}}
\end{figure*}
